No Replication Without Process Documentation:  An Assessment of Management

### 2020-05-07 R/MSE-manuscript.Rmd
Update to refect text edits in manuscript submission

---
title: 'No Replication Without Process Documentation:  An Assessment of Management
  Strategy Evaluations'
author:
- affiliation: School for Marine Science and Technology, University of Massachusetts
    - Dartmouth
  name: Jonathan W. Cummings, Amanda Hart, Gavin Fay
output:
  bookdown::word_document2:
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
    number_sections: false
always_allow_html: yes
biblio-style: apalike
endnote: no
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
header-includes: \usepackage{hyperref}
keywords: pandoc, r markdown, knitr
bibliography: MSEreview.bib
abstract: The aim and vision for fisheries management that comes from management strategy evaluation and adaptive management is a noble and aspirational vision.  We review MSE documentation in the published literature to appraise the achievement of this vision, finding that reality doesn't fully reflect this vision. The steps taken to complete the management strategy evaluation process often are not documented. Scientists often filled the role of fishery stakeholders, seemingly selecting the objectives and alternatives to evaluate in most management strategy evaluations. Facilitators and decision analysts -- participants focused on the decision making process rather than making the decision -- were rarely participants in MSE processes.  Conservation, Economic, and Yield focused objectives were commonly considered in MSE processes, while broader social objectives were rarely considered.  The picture painted by MSE documentation is primarily one of scientist driven simulation studies rather than management driven decision processes, with limited process documentation occurring. We hope providing a repository for reviews of previous management strategy evaluations will aid and speed learning among practitioners of management strategy evaluation.
---

@@

# Introduction 
The calls for fisheries management to utilize management strategy evaluation and adaptive management have been both aspirational and estimable in their aim and vision. For example, management strategy evaluation (MSE) is "widely considered to be the most appropriate way to evaluate the trade‐offs achieved by alternative management strategies and to assess the consequences of uncertainty for achieving management goals” [@Punt]. The positive aspects of MSE that have been listed include that it is “A tool that scientists and managers can use to simulate the workings of a fisheries system and allow them to test whether potential harvest strategies — or management procedures — can achieve pre-agreed management objectives.”[@Pew], that it “can be used to identify a ‘best’ management strategy among a set of candidate strategies" [@PuntLab] and “MSE has the advantage of being able to reveal the trade-offs among a range of possible management decisions, and should assist our Commissioners in determining whether the objectives should be weighted differently. Specifically, to provide the information on which to base a rational decision, given objectives, preferences, and attitudes to risk.”  [@IPHC].  Based on these descriptions MSE is a compelling tool to support and improve fisheries management decisions.

Are the results of MSE living up to their descriptions? Are MSEs evaluating trade-offs and the consequences of uncertainty? Do they test potential harvest strategies? Were the management objectives used to evaluate management procedures established and agreed to? Was a 'best' management strategy identified? Was consideration given to the weights on objectives, preferences, and attitudes to risk?

In the descriptions of the MSE process, the process itself is described as being "at the interface between science and policy.” [@PuntLab] and being an “Undertaking [that] requires scientists, managers, and stakeholders to be involved throughout the process." [@IPHC] The International Pacific Halibut Commission (2017) also states, "While the scientists do the modelling, managers must offer extensive input. Because of the many steps and the iterative process, communication among parties is critical for achieving buy-in on the results of the management strategy evaluation.”

Are MSE processes being conducted in the manner described? Do MSE processes consider and account for the interface between science and policy? Are scientists, managers and stakeholders involved and communicating through and iterative process that creates buy-in?

Adaptive management developed in fisheries management as a means to improve the achievement of objectives through time, when learning can aid achieving those objectives [@ADM].  For example, the Millennium Ecosystem Assessment [-@MEA] defines adaptive management as “A systematic process for continually improving management policies and practices by learning from the outcomes of previously employed policies and practices. In active adaptive management, management is treated as a deliberate experiment for purposes of learning.” We contend that scientific publication is essentially a process intended to support adaptive management, in that researchers gain knowledge from the decisions made and outcomes produced by previous work to inform their decisions. In the case of fisheries management, are scientific publications of management strategy evaluations supporting adaptive management?

## Objectives 
Our overall objective is to determine the extent that MSE achieves the aspirational aims outlined above, and whether MSE documentation supports learning within the MSE practitioner community.  Given that scientific publication is the primary means of communicating the results of MSE, we review the scientific literature to assess the extent to which the process and aims of MSE, and adaptive management are being achieved and supported by the scientific literature. To assess this, we determine how MSEs utilize standard components of decision making processes. We use the structured decision making (SDM) process [@SDM] -- the decision making framework in which adaptive management occurs -- as our framework for evaluation.  Using this framework, we review published MSEs to:

* Determine what questions MSE processes are addressing
* Identify the objectives and alternatives considered in MSE
* Determine the prevalence of a) explicit documentation and b) use of decision making steps in the MSE process, and 
* Identify participants in MSE processes.

An increasingly recognized challenge in fisheries management is responding to changes in the productivity of fisheries stocks as a result of climate change [@busch_climate_2016; @brander_global_2007]. We also conduct a focused analysis of MSE processes that address the influence of climate change on fisheries management, highlighting aspects of MSE processes that model climate change as a driver of system status.

# Methods 
## Finding and Sampling the MSE Literature 
We conducted our search for MSEs in the SCI-EXPANDED index from Web of Science, searching for “management strategy evaluation” by topic across all years on January 8th, 2019 (Table \@ref(tab:MSEcount)). This search returned 264 results.  We reviewed a random sample of 30 articles that document a MSE (Appendix B), removing articles that were reviews, meta-analyses, or simply cited other MSE articles from our sample. After removing other articles from the original set of 264 articles, 154 articles document a MSE. The number of MSEs published each year is trending higher (Figure \@ref(fig:YearHistogram)), with the ICES Journal of Marine Science and Fisheries Research as the leading publication outlets (Figure \@ref(fig:PublicationHistogram)).

@@

We reviewed 16 climate change MSE articles (Appendix B). Twelve of these articles came from our original January 8th, 2019 Web of Science search.  To give as full a picture of the climate change MSE articles to date as possible we also reviewed the 4 climate change MSE articles published in 2019.

@@

@@

We present the review methodology and results in four groups: consequences - which focuses on what was considered during the prediction stage, objectives - which focuses on what categories of objectives were considered, stages - which focuses on whether and how the stages of the process were completed, and participants - which focuses on how participation in the process was structured.  These results were entered into a relational database for storage and evaluation, and is available via our shiny application (https://jonathancummings.shinyapps.io/MSEreview/).

The consequence group addresses what question the MSE was designed to address, and how it was addressed. We report the drivers included as the primary factors influencing the predictions produced by the MSE model, listing each driver a MSE included. We also report how the consequence predictions were made, with possible values of simulation modeling, dynamic programming, expert elicitation, and mental models, or unknown when it is unclear what methodology was used to predict the status of the system for management evaluation.

In the objectives group we address what the concerns, or goals of the MSE were.  We record each objective considered, and report the type and category of objectives included in each MSE. Natural resource management problems typically include four types of objectives: conservation, resource use, i.e., yield in the fisheries context, as well as economic, and social objectives [@runge2013]. We sort objectives into four types: strategic objectives - which are concerned with the overall mission of the system in which the decision evaluated by the MSE falls, process objectives - which are concerned with how the management decision addressed by the MSE would be made, fundamental objectives - which address the goals of the management decision, and means objectives - which address means to achieving the fundamental objectives of the management decision.

In the stages group we address how the MSE stages were completed. We report whether the stages of the structured decision making framework, i.e., problem, objectives, alternatives, tradeoffs, were explicitly completed and the methods used to complete them were documented.  We also report whether the MSE results were adopted in subsequent management.

In the participant group we address how participation in the MSE process was structured and who participated. We report who lead the MSE process, who participated in it, and if the roles of participants were established and documented.  In many cases the process used to elicit objectives and alternatives was not explicit.  In these cases, we used what information was provided to glean who participated in the production of those components of the MSE.  We record and report who those participants were, referring to them as subjective participants in the objectives or alternatives process.  In other cases where there was not enough information to glean who participated we report the participants as unknown.

We collected additional information about the MSEs we reviewed beyond what is reported here. For a full description of the information recorded in the relational database see the supplemental information (Appendix A), and for the full data set and results see our shiny application (https://jonathancummings.shinyapps.io/MSEreview/).

# Results 
## What questions and management objectives do MSEs address? 
We found that uncertainty was the main driver of the system state included in management strategy evaluations, with uncertainty included in 21 of the 30 evaluations we reviewed in our random sample (Table \@ref(tab:drivers)).  In addition to uncertainty about the system model itself, monitoring methodology, how and what data is collected, was the most prevalent driver in MSE models, but a variety of drivers were considered across the random sample of MSEs.

```{r drivers}
drive.data_pub %>%
  mutate("Percent"=round(Percent,0)) %>% 
  select(Driver,Percent,Frequency) %>% 
  kable(caption = "Drivers included in management strategy evaluations") %>% 
  kable_styling()
```
<br>
Conservation objectives were present in nearly all of the MSEs we sampled (Table \@ref(tab:objectiveCategories)).  Yield and economic objectives were present in about three quarters of the MSEs, while social objectives were only included in 2 of the sampled MSEs.

```{r, objectiveCategories}
objcat.data_pub %>% 
  select(-Analysis) %>% 
  kable(caption = "Consideration of objective categories in management strategy evaluation") %>% 
  kable_styling()
```
<br>
No strategic or process objectives were documented in the MSEs.  There were roughly the same number of fundamental and means objectives included in the MSEs (Table \@ref(tab:objectiveTypes)). An average of 6 objectives were evaluated by each management strategy evaluation.
```{r, objectiveTypes}
obj.dataTable_pub %>%
  filter(Objective=="Type") %>%
  select(Type, Percent, Number, 'Per MSE', -Analysis, -Objective) %>% 
  kable(caption = "Objective Types") %>% 
  kable_styling()
```

## Are structured decision making steps explicit in MSEs? 
The majority of MSEs did not explicitly document how the stages of the MSE process -- defined using the SDM framework -- were completed (Table \@ref(tab:explicit)).  Apart from the consequence stage, which is the primary focus of MSE publication and therefore was always addressed, the most documented stage was the alternative stage with 6 MSEs documenting this stage of the process. The MSE process, problem, and objectives methodology were each documented in 5 of the MSEs. Three of the 30 MSEs explicitly documented a trade-off analysis and 2 of the 30 documented a management decision. One MSE documented that the results of the MSE were adopted.

@@

## Who is involved in MSEs? 
MSEs were rarely explicit about the roles participants played in MSE processes, as only one MSE documented the roles participants played in the process (Table \@ref(tab:roles)).  None of the MSEs noted whether or not public meetings were held in which the public was invited to participate.

@@

The primary participants and leaders of the MSE processes were scientists, while some MSE process were led by governments and management agencies (Figure \@ref(fig:participants) Process).  Scientists were explicitly the leaders -- or the sole participants and therefore seemingly the leaders -- in 73 percent of the MSE processes, while 27 percent of processes lacked enough documentation of participants and leadership such that we were unable to identify a presumed leader of the process.

Scientists were participants in nearly all of the reviewed MSE processes (Figure \@ref(fig:participants) Participants).  Other stakeholders in MSE processes were less frequent participants.  Members of the fishery participated in 30 percent, management and government representatives in 23 percent, representatives of independent institutions in 7 percent, and the members of the public in 3 percent of reviewed MSEs.  Decision makers -- those responsible for selecting the management plan to implement -- were participants in 13 percent of MSEs.  Experts -- a possible source of data or predictions -- participated in 3 percent of the MSEs reviewed.  Participants involved to the assist the MSE process itself, i.e., facilitators and decision analysts, participated in 10 and 3 percent of MSE processes respectively.

Objectives and alternatives were elicited from a variety of participants. Management, government, or fishery participants explicitly provided objectives and alternatives in about 10 percent of the MSE processes (Figure \@ref(fig:participants) Explicit).  However, in most cases the source of the objectives and alternatives was not explicit.  The source of objectives and alternatives was not explicit in more than 80% of the MSE processes, 25 of 30 objective sources were not explicit and 24 of 30 alternative sources. In those subjective cases scientists -- being the only or primary participants -- seemingly selected these components most frequently (Figure \@ref(fig:participants) Subjective).

@@

# Climate Change in MSE (As *Box Insert*) 
Fisheries management has mostly focused on fishing impacts with ecosystem status viewed as a background constant. The increasing rate of climate change and its influence on fisheries is changing this dynamic, bringing ecosystem status to the forefront of fisheries management.

Here we review published MSEs the include climate change as a driver of the fisheries model using the stages in a SDM process to highlight unique aspects of these MSEs. We filtered the original set of MSEs published by as of December 31st, 2019 searching for the term "climate change" within the set that resulted from searching for "management strategy evaluation". This resulted in our reviewing 16 published MSEs that included climate change as a driver. We reviewed the climate change MSEs using the same methodology described above. These MSEs have been concentrated in North America and Australia (Figure \@ref(fig:mapfigure)).

As with the random sample of reviewed MSEs, uncertainty was the most common driver, apart from climate change, included climate change MSEs (Table \@ref(tab:driversCC)), while other biological drivers were also included in some of the climate change MSEs.  Overall climate change MSEs considered fewer drivers on average, 1.4, than the random sample of MSEs, which included 1.8.

@@

Climate change MSEs averaged the same number of objectives per MSE. The categories of the objectives differed however, with more frequent inclusion of yield objectives, but less frequent consideration of economic objectives and no consideration of social objectives (Table \@ref(tab:CCobjectiveCategories)) in the climate change MSEs.
```{r, CCPerMSE}
# color_Pub <- which(per.MSE$Analysis=='Random Sample')
# color_CC <- which(per.MSE$Analysis=='Climate Change')
# 
# per.MSE %>%
#   kable(caption = "Number of objectives considered per management strategy evaluation") %>%
#   row_spec(color_CC, color = "#66CCCC", background = "white") %>%
#   row_spec(color_Pub, color = "#006666", background = "white")%>% 
#   kable_styling()
```

```{r, CCobjectiveCategories}
objcat.data_CC<-objcat.data %>%
  filter(`Objective Category`!="Utility") %>% 
  add_column(order=c(1,2,5,6,7,3,4)) %>% 
  arrange(order)

color_Pub <- which(objcat.data_CC$Analysis=='Random Sample')
color_CC <- which(objcat.data_CC$Analysis=='Climate Change')

objcat.data_CC %>%
  select(-order) %>% 
  kable(caption = "Objective categories considered in management strategy evaluations") %>%
  row_spec(color_CC, color = "#66CCCC", background = "white") %>%
  row_spec(color_Pub, color = "#006666", background = "white") %>% 
  kable_styling()
```

Climate change MSEs were more likely to explicitly document the trade-off analysis stage, equally likely to document the alternatives stages, but were less likely to document the other stages in the decision process (Table \@ref(tab:CCstages)).
```{r, CCstages}
freq.data_CC %>%
  mutate("Percent"=round(Percent,0)) %>% 
  select(Explicit, Percent, Number, -Analysis) %>% 
  kable(caption = "Prevalence of explicit documentation of stages in climate change MSE processes") %>%
  kable_styling()
```

Including climate change in a MSE appears to reduce the complexity of other stages in the MSE process, in particular those stages that utilize input from additional participants. On average climate change MSEs consider fewer drivers, consider less diverse objectives, and explicitly document fewer stages of the MSE process.  If the MSE process is time, effort, or participant limited the inclusion of climate change may be utilizing time, effort, or outside participants to focus on climate change rather than other aspects of the MSE process.

# Discussion 
The primary picture painted by MSE documentation is one of scientist driven simulation studies rather than management driven decision processes. While documentation about the modeling conducted in a MSE was explicit, documentation about the MSE process itself was often limited.

We found that management strategy evaluations do typically assess the consequences of uncertainty for achieving management goals, but other factors influencing the consequences of management strategies get more sporadic treatment.  Data limitation is a frequent concern in fisheries management, and contributes to the degree of uncertainty present in fisheries management.  Next to accounting for uncertainty, accessing the influence of data quality and availability, i.e.,  monitoring methodology was the most prevalent driver included in MSEs.  This likely points to a desire for the design and use of cost-effective fisheries monitoring plans, suggesting future MSEs may be able to learn from the set including monitoring methodology as a driver. Additionally, the use of value of information analyses could expand upon the benefits of MSE as a tool to direct scientific resources [@runge_2011; @moore_2012]. While discussions of fisheries research and management seem to be moving toward ecosystem effects, spatial models, and multiple species models, these characteristics were rarely considered in the set of MSEs we reviewed.

Managing natural resources entails achieving the objectives of people interacting with a natural resource, so decision processes are likely to produce more desirable results when they include objectives that reflect stakeholders' and society's values.  These objectives are likely to include conservation, yield, economic, and social concerns. As noted, MSEs are described as a tool to evaluate what management strategy best achieves objectives, however the MSEs we reviewed don't cover all categories of objectives. Yield and economic objectives were not included in about one quarter of MSEs, and broader social objectives were absent from all but 2 of the 30 evaluations. While we are unable to know the cause for this without more explicit documentation of MSE processes, our results may provide some clues.

>>>>

---
title: 'No Replication Without Process Documentation:  An Assessment of Management
  Strategy Evaluations'
author:
- affiliation: School for Marine Science and Technology, University of Massachusetts
    - Dartmouth
  name: Jonathan W. Cummings, Amanda Hart, Gavin Fay
output:
  bookdown::word_document2:
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
    number_sections: false
always_allow_html: yes
biblio-style: apalike
endnote: no
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
header-includes: \usepackage{hyperref}
keywords: pandoc, r markdown, knitr
bibliography: MSEreview.bib
abstract: The aim and vision for fisheries management that comes from management strategy evaluation and adaptive management is a noble and aspirational vision. We review MSE documentation in the published literature to appraise the achievement of this vision, finding that reality doesn’t fully reflect this vision. As a scientific research tool to evaluate management performance, exemplary applications of MSE in the literature abound. Examples of full integration of the MSE process as a stakeholder-based participatory, management decision support framework are less common in the literature, and these processes are not as well documented as the technical modeling detail. The steps taken to complete the management strategy evaluation process often are not documented and scientists often filled the role of fishery stakeholders, seemingly selecting the objectives and alternatives to evaluate in most management strategy evaluations. Facilitators and decision analysts – participants who may aid the process by focusing on the decision-making process rather than making the decision – were rarely participants. Conservation, Economic, and Yield focused objectives were commonly considered in MSE processes, while broader social objectives were rarely considered. While the picture painted by MSE documentation is primarily one of scientist driven simulation studies rather than management driven decision processes, with limited process documentation occurring. Here we provide a repository for reviews of previous management strategy evaluations with hopes this will aid and speed learning among practitioners of management strategy evaluation.
---

@@

# Introduction 
The calls for fisheries management to use management strategy evaluation and adaptive management have been both aspirational and estimable in their aim and vision. For example, management strategy evaluation (MSE) is “widely considered to be the most appropriate way to evaluate the trade‐offs achieved by alternative management strategies and to assess the consequences of uncertainty for achieving management goals” [@Punt]. The positive aspects of MSE that have been listed include that it is “A tool that scientists and managers can use to simulate the workings of a fisheries system and allow them to test whether potential harvest strategies — or management procedures — can achieve pre-agreed management objectives.”[@Pew], that it “can be used to identify a ‘best’ management strategy among a set of candidate strategies” [@PuntLab] and “MSE has the advantage of being able to reveal the trade-offs among a range of possible management decisions, and should assist our Commissioners in determining whether the objectives should be weighted differently. Specifically, to provide the information on which to base a rational decision, given objectives, preferences, and attitudes to risk.” [@IPHC]. Based on these descriptions MSE is a compelling tool to support and improve fisheries management decisions.

Are MSEs meeting the aspirational standards for aiding management-driven processes as outlined in the literature?” Do MSEs evaluate trade-offs and the consequences of uncertainty? Do they test potential harvest strategies? Were the management objectives used to evaluate management procedures established and agreed to? Was a 'best' management strategy identified? Was consideration given to the weights on objectives, preferences, and attitudes to risk? These questions prompted the review that follows.
Practitioners of MSE have described it as a method that lies “at the interface between science and policy” [@PuntLab] and an “Undertaking [that] requires scientists, managers, and stakeholders to be involved throughout the process” [@IPHC] also states, that “While the scientists do the modelling, managers must offer extensive input. Because of the many steps and the iterative process, communication among parties is critical for achieving buy-in on the results of the management strategy evaluation.”

These statements raise additional questions for our review. Are MSE processes being conducted in the manner described? Do MSE processes consider and account for the interface between science and policy? Are scientists, managers and stakeholders involved and communicating through and iterative process that creates buy-in?

Similar to MSE, adaptive management was developed in fisheries management as a means to improve the achievement of objectives through time, when learning can aid achieving those objectives [@ADM].  For example, the Millennium Ecosystem Assessment [-@MEA] defines adaptive management as “A systematic process for continually improving management policies and practices by learning from the outcomes of previously employed policies and practices. In active adaptive management, management is treated as a deliberate experiment for purposes of learning.” We contend that scientific publication is essentially a process intended to support adaptive management, in that researchers gain knowledge from the decisions made and outcomes produced by previous work to inform their decisions. This led to our review’s final question, are scientific publications of management strategy evaluations supporting adaptive management?

## Objectives 
Our overall objective for this review was to determine the extent to which MSE achieves the aspirational aims outlined above and evaluate whether MSE documentation supports learning within the MSE practitioner community. Given that scientific publication is the primary means of communicating the results of MSE, we reviewed the scientific literature to assess this objective. We evaluated whether MSEs use standard components of decision making processes using the structured decision making (SDM) process [@SDM] -- the decision making framework in which adaptive management occurs -- as our framework for evaluation.  Using SDM as our framework, we:

* Determined what questions MSE processes are addressing
* Identified the objectives and alternatives considered in MSE
* Determined the prevalence of a) explicit documentation and b) use of decision making steps in the MSE process, and 
* Identified participants in MSE processes.

As MSE expands and seeks to address new challenges, the insights from past MSEs may become increasingly valuable. Fisheries management has mostly focused on fishing impacts with ecosystem status viewed as a background constant. The increasing rate of climate change and its influence on fisheries is changing this dynamic, bringing ecosystem status to the forefront of fisheries management [@busch_climate_2016; @brander_global_2007]. We use climate change focused MSEs as a case study into the use of MSE in an emerging area of study and conduct a focused analysis of MSE processes that address the influence of climate change on fisheries management, highlighting aspects of MSE processes that model climate change as a driver of system status.

# Methods 
## Finding and Sampling the MSE Literature 
We conducted our search for MSEs in the SCI-EXPANDED index from Web of Science, searching for “management strategy evaluation” by topic across all years on January 8th, 2019 (Table \@ref(tab:MSEcount)). This search returned 264 results. We reviewed a random sample of 30 articles that document a MSE (Appendix A), removing articles that were reviews, meta-analyses, or simply cited other MSE articles from our sample. After removing other articles from the original set of 264 articles, 154 articles document a MSE. Searches for MSEs by other names i.e., “management procedure approach” and “management procedure evaluation” returned 9 and 1 results respectively, none of which document a MSE. The number of MSEs published each year is trending higher (Figure \@ref(fig:YearHistogram)), with the ICES Journal of Marine Science and Fisheries Research as the leading publication outlets (Figure \@ref(fig:PublicationHistogram)).

@@

@@

We reviewed 16 climate change MSE articles (Appendix A). Twelve of these articles came from our original January 8th, 2019 Web of Science search. To give as full a picture of the climate change MSE articles to date as possible we also reviewed the 4 climate change MSE articles published in 2019. These MSEs have been concentrated in North America and Australia (Figure \@ref(fig:mapfigure)).

@@

We present the review methodology and results in four groups: stages - which focuses on whether and how the stages of the process were completed, objectives - which focuses on what categories of objectives were considered, participants - which focuses on how participation in the process was structured, and consequences - which focuses on what was considered during the prediction stage. These results were entered into a relational database for storage and evaluation, and are available via our shiny application (https://jonathancummings.shinyapps.io/MSEreview/).

In the stages group we identify how the MSE stages were completed. We report whether the stages of the structured decision making framework, i.e., problem, objectives, alternatives, tradeoffs, were explicitly completed and the methods used to complete them were documented. We also report whether the MSE results were adopted in subsequent management.

In the objectives group we identify the concerns, or goals of each MSE. We record each objective considered, and report the type and category of objectives included in each MSE. Natural resource management problems typically include four categories of objectives: conservation, resource use, i.e., yield in the fisheries context, as well as economic, and social objectives [@runge2013]. At times multiple objectives categories are combined into a single metric, which we considered a fifth objective category, namely utility objectives. We sort objectives into four types: 1) strategic objectives - which are related to the overall mission of management in the system where the decision evaluated by the MSE occurs, e.g., objectives reaching beyond the scope of the problem addressed, 2) process objectives - which are concerned with how the management decision addressed by the MSE would be made, e.g., achieving stakeholder buy-in to project outcomes, 3) fundamental objectives - which address the goals of the management decision, e.g., maximizing revenue from the fishery, and 4) means objectives - which address means to achieving the fundamental objectives of the management decision, e.g. maximizing catch per unit effort, which is a means to maximizing revenue.

In the participant group we describe how participation in each MSE process was structured and who participated. We report who led the MSE process, who participated in it, and if the roles of participants were established and documented. In many cases the process used to elicit objectives and alternatives was not explicit. In these cases, we used what information was provided to glean who participated in the production of those components of the MSE. We record and report who those participants were, referring to them as subjective participants in the objectives or alternatives process. In cases where there was not enough information to glean who participated we report the participants as unknown.

The consequence group identifies what question each MSE was designed to address, and what methods were employed. We report the drivers – the characteristics of the model that drive the behavior of and determine the predictions produced by the MSE model – listing each driver a given MSE included. We also report whether consequence predictions were made, using simulation modeling, dynamic programming, expert elicitation, mental models, or unknown prediction methods when it was unclear what methodology was used to predict the status of the system for management evaluation.
We collected additional information about the MSEs we reviewed beyond what is reported here. For a full description of the information recorded in the relational database see our appendix (Appendix B), and for the full data set and results see our web-based shiny application in R  (https://jonathancummings.shinyapps.io/MSEreview/).

# Results 
## Are structured decision making steps explicit in MSEs? 
The majority of MSEs did not explicitly document how the stages of the MSE process -- defined using the SDM framework (Figure \@ref(fig:SDMfigure)) -- were completed (Table \@ref(tab:explicit), Random Sample).  Apart from the consequence stage, which is the primary focus of MSE publication and therefore was always addressed, the most documented stage was the alternative stage with 6 MSEs documenting this stage of the process. The MSE process, problem, and objectives methodology were each documented in 5 of the MSEs. Three of the 30 MSEs explicitly documented a trade-off analysis and 2 of the 30 documented a management decision. One MSE documented that the results of the MSE were adopted. Climate change MSEs were more likely to explicitly document the trade-off analysis stage, equally likely to document the alternatives stages, but less likely to document the other stages in the decision process (Table \@ref(tab:explicit), Climate Change).

@@

## What management objectives do MSEs address? 
We found that uncertainty was the main driver of the system state included in management strategy evaluations, with uncertainty included in 21 of the 30 evaluations we reviewed in our random sample (Table \@ref(tab:drivers)).  In addition to uncertainty about the system model itself, monitoring methodology, how and what data is collected, was the most prevalent driver in MSE models, but a variety of drivers were considered across the random sample of MSEs.

```{r drivers}
drive.data_pub %>%
  mutate("Percent"=round(Percent,0)) %>% 
  select(Driver,Percent,Frequency) %>% 
  kable(caption = "Drivers included in management strategy evaluations") %>% 
  kable_styling()
```
<br>
Conservation objectives were present in nearly all of the MSEs we sampled (Table \@ref(tab:objectiveCategories)). Yield and economic objectives were present in about three quarters of the MSEs, while social objectives were only included in 2 of the sampled MSEs. The climate change MSE objectives differed, with more frequent inclusion of yield objectives, but less frequent consideration of economic objectives and no consideration of social objectives.

```{r, objectiveCategories}
objcat.data_pub %>% 
  select(-Analysis) %>% 
  kable(caption = "Consideration of objective categories in management strategy evaluation") %>% 
  kable_styling()
```
<br>
No strategic or process objectives were documented in the sampled MSEs. There were roughly the same number of fundamental and means objectives included in the MSEs (Table \@ref(tab:objectiveTypes)). An average of 6 objectives were evaluated by each management strategy evaluation. On average climate change MSEs included the same number of objectives per MSE as the random sample.
```{r, objectiveTypes}
obj.dataTable_pub %>%
  filter(Objective=="Type") %>%
  select(Type, Percent, Number, 'Per MSE', -Analysis, -Objective) %>% 
  kable(caption = "Objective Types") %>% 
  kable_styling()
```

## Who is involved in MSEs? 
MSEs were rarely explicit about the roles participants played in MSE processes, with only one MSE documenting these (Table \@ref(tab:roles)). None of the MSEs noted whether or not public meetings were held in which the public was invited to participate.

@@

Inferring from scientifically authored publications in the absence of explicit statements indicating otherwise, the primary participants and leaders of the MSE processes were scientists, while some MSE processes were led by governments and management agencies (Figure \@ref(fig:participants) Process).  Scientists were explicitly the leaders – or the sole participants and therefore seemingly the primary participants and leaders – in 73 percent of the MSE processes, while 27 percent of processes lacked enough documentation of participants and leadership such that we were unable to identify a presumed leader of the process.

Scientists were participants in all of the reviewed MSE processes (Figure \@ref(fig:participants) Participants).  Other stakeholders in MSE processes were less frequent participants. Members of the fishing community participated in 30 percent, management and government representatives in 23 percent, representatives of independent institutions in 7 percent, and the members of the public in 3 percent of reviewed MSEs. Decision makers – those responsible for selecting the management plan to implement – were participants in 13 percent of MSEs. Experts – a possible source of data or predictions – participated in 3 percent of the MSEs reviewed. Participants involved to assist the MSE process itself, i.e., facilitators and decision analysts, participated in 10 and 3 percent of MSE processes respectively.

Objectives and alternatives were elicited from a variety of participants. Management, government, or fishery participants explicitly provided objectives and alternatives in about 10 percent of the MSE processes (Figure \@ref(fig:participants) Explicit).  However, in most cases the source of the objectives and alternatives was not explicitly documented. The source of objectives and alternatives was not explicit in more than 80% of the MSE processes, 25 of 30 MSE processes used objectives without explicitly documenting the source for the objectives, i.e., who provided them, and 24 of 30 MSEs did not explicitly document who provided the alternatives. In those subjective cases scientists – being the only or primary participants – seemingly selected these components, making them the most frequent providers of objectives and alternatives in MSE processes (Figure \@ref(fig:participants) Subjective).

@@

## What questions do MSEs address? 
All sampled MSEs used simulation modeling for consequence prediction. We found that uncertainty was the main driver of the system state included in management strategy evaluations, with uncertainty included in 21 of the 30 evaluations we reviewed (Table \@ref(tab:driversCC)). Monitoring methodology, how and what data was collected, was the next most prevalent driver in MSE models, but a variety of drivers were considered across the random sample of MSEs. As with the random sample of reviewed MSEs, uncertainty was the most common driver, apart from climate change, included in climate change MSEs, while other biological drivers were also included in some of the climate change MSEs. Overall, climate change MSEs considered fewer drivers on average (1.4) than the random sample of MSEs (1.8).

@@

# Discussion 
The primary picture painted by MSE documentation in the scientific literature is one of scientist driven simulation studies rather than management driven decision processes. Documentation about the modeling conducted in a MSE was explicit, and all MSEs included well documented simulation studies supporting decision making based upon a clear prediction of the future state of the selected performance metrics. Documentation about the MSE process itself was often limited, however.

We found that management strategy evaluations typically assess the consequences of uncertainty for achieving management goals, one of the recommended best practices of MSEs [@Punt], but other factors influencing the consequences of management strategies get more sporadic treatment. Data limitation is a frequent concern in fisheries management and contributes to the degree of uncertainty present in fisheries management. Next to accounting for uncertainty, accessing the influence of data quality and availability, i.e., monitoring methodology was the most prevalent driver included in MSEs. This likely points to a desire for the design and use of cost-effective fisheries monitoring plans, suggesting future MSEs may be able to learn from the set of studies identified here that include monitoring methodology as a driver. Additionally, using value of information analyses – which directly determine the expected benefits of additional research – could expand upon the benefits of MSE as a tool to direct scientific resources [@runge_2011; @moore_2012]. While discussions of fisheries research and management seem to be moving toward ecosystem effects, spatial models, and multiple species models, these characteristics were rarely considered in the set of MSEs we reviewed.
Including climate change in a MSE appears to reduce the complexity of other stages in the MSE process, in particular those stages that use input from additional participants. On average climate change MSEs consider fewer drivers, consider less diverse objectives, and explicitly document fewer stages of the MSE process. If the MSE process is time, effort, or participant limited, the inclusion of climate change may be utilizing time, effort, or outside participants to focus on climate change rather than other aspects of the MSE process.

Managing natural resources entails achieving the objectives of people interacting with a natural resource, so decision processes are likely to produce more desirable results when they include objectives that reflect stakeholders’ and society’s values. These objectives are likely to include conservation, yield, economic, and social concerns. As noted, MSEs are described as a tool to evaluate which management strategies best achieve objectives, however the MSEs we reviewed don’t cover all categories of objectives. Yield and economic objectives were not included in about one quarter of MSEs, and broader social objectives were absent from all but 2 of the 30 evaluations. While we are unable to know the cause for this without more explicit documentation of MSE processes, our results may provide some clues.

Scientists led, and participated in most of the MSEs, while decision-makers, decision making institutions, and stakeholders were relatively infrequent leaders or participants in the process. The lack of participation by stakeholders may result in less frequent inclusion of social objectives, as well as yield and economic objectives specific to a particular fishery. Another factor may be that MSE models are likely constructed mainly by ecological modelers, who are likely to be more comfortable modeling the consequences of conservation and yield objectives than economic or social objectives. Given that scientists seem to be providing the objectives in many MSE processes, they may be selecting objectives that they are more comfortable evaluating. There also appears to be some degree of borrowing occurring in which scientists use a set of generic fishery objectives when conducting a MSE. Perhaps because there were few social objectives included in these MSEs, there was little inclusion of experts to aid in the prediction of consequences, or consequence prediction techniques other than simulation modeling.

Participants that would likely expand the set of objectives considered in a MSE rarely participated in the MSEs we reviewed. That is, facilitators and decision analysts were rarely participants in MSE processes. The inclusion of facilitators in decision processes aid information gain and conflict resolution, increase trust between participants, and lead to more beneficial environmental and social outcomes [@Vente; @Feeney2019].  The additional structure and focus on the decision process would likely result in more explicit documentation when facilitators and decision analysts participate in MSE, as resulted in some of the MSEs reviewed here [@williams2011; @smith2013; @kolody2008]. Facilitators and decision analysts can also aid decision processes by helping to diagnose the problem, which can result in selecting more appropriate analysis techniques for the problem type, elicit a more complete and representative set of objectives, and select analysis tools to best address decision impediments to identify the optimal management strategy [@Cummings_inprep].

The MSE process may be aided by utilizing the techniques from the structured decision making approach to decision analysis, such as expanding the focus on problem framing at the outset of MSEs, eliciting fully representative sets of objectives, and utilizing multiple attribute utility theory to evaluate tradeoffs between objectives [@Cummings_inprep], and more closely adhering to the best practices for MSE [@Punt]. Utilizing the structured decision making approach also eases the documentation process by breaking decision making into more discrete stages. However, there may be practical considerations, such as time, budgetary constraints, or available participants that limit how complex the stages of a MSE are, and the degree or form of documentation that occurs. Where the expertise and budget are available, the use of facilitators and scribes may aid in addressing and documenting MSE processes.

We did not include grey literature in our evaluation. It is possible that more of the MSE process is documented in grey literature publications, and less is documented in the scientific publications because they are primarily produced by scientific authors. However, we are aware of only one grey literature publication focused on aspects of an MSE process, namely grey literature documentation of the Atlantic herring MSE process. We also note that grey literature publications are less likely to reach an audience of MSE practitioners and perhaps should be included as an appendix or supplement to MSE publications where available.
